Summer 2025 ToDo

### 2025-10-24 Relations/Paper/Well-foundedness.tex
Minor fixes to WF

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.  This definition appears to be strictly weaker than
WFseqm-not-not, asserting that every sequence not-not contains a
non-decreasing index.
\apcomment{fix the macro please!}
% \apcomment{Perhaps we should have a separate discussion between isWFseq- and isWFseq-not-not?}

@@

@@

{
\def\arraystretch{1.3}
\begin{figure}[h!]
\small
\begin{tabular}{@{}l l l @{}}
\toprule
\textbf{Property} & \textbf{Definition}  &\textbf{Agda Code}  \\
\midrule
$\rdec$   & $R$ is decidable
          & $\forall$ \verb|{x} {y}| $\to$ \verb|EM (R x y)|\\
$\FB$     & $R$ is finitely branching
          & $\mathtt{\forall (a : A) \to \Sigma [ xs \in List A ] (\forall b \to R a b \to b \in{}List\, xs)}$ \\
$\mpe$    & Sequences are $\lnot\lnot$-closed
          & $\forall\ \mathtt{(f : \nat \to A) \to \lnot\lnot Closed
          (\lambda x \to \Sigma [ k \in \nat ] (f\, k \equiv x))}$ \\
$\corDNE$ & Coreductives are $\lnot\lnot$-closed
          & $\mathtt{\forall (P : \mathcal{P} A) \to R}$-$\mathtt{coreductive\ P \to \lnot\lnot Closed\ P}$\\
$\accDNE$ & Accessibility is $\lnot\lnot$-closed
          & $\lnot\lnot$\verb|Closed (R -accessible)| \\
$\accCor$ & Accessibility is coreductive
          & \verb|R -coreductive (R -accessible)| \\
\bottomrule
\end{tabular}
\centering
\caption{Classical Properties Used in the Well-Foundedness Diagram}
\label{tab:cprop}
\end{figure}
}

% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.
Figure \ref{fig:WF} illustrates that the network of logical relationships between
constructive notions of well-foundedness is quite rich indeed.
Taking the $\lnot\lnot$-closures of
these notions cuts down some, but not all, of the complexity.
For example, the two minimality principles become equivalent.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the weak version of the sequential definition is unable to imply
the (weak) accessible one without an additional hypothesis $\accCor$,
which may be described as the constructive contrapositive of the definition of accessibility.
In the context of rewriting systems,
this condition asserts the existence of a function which maps any
element that is \emph{not} strongly normalizing to a one-step reduct of it having the same property.
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

Other observations that can be gleaned from the figure include the following.
\apcomment{The macros have to fixed throughout.}
\begin{enumerate}
  \item The implication from $\isWFacc$ to $\isWFseq$ requires the underlying relation to be decidable.
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires the image of every sequence to be $\lnot\lnot$-closed
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (if one assumes that equality on $A$ is decidable).
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent, are directly implied by $\isWFaccm$,
    and directly imply $\isWFseqm$  \apcomment{Fix macros}.
  \item All definitions imply $\WFcorm$.  This weakest definition implies $\isWFaccm$ assuming the
    ``constructive contrapositive of accessibility'', $\accCor$.
    %While this condition is implied by $\isWFseqm$, it does not appear to imply anything on its own.
  \item Assuming $\accCor$, $\isWFacc$ can also be derived from $\WFcor$. % , the stronger version of $\WFcorm$.
    \apcomment{Remove the ``is''!}

    \apcomment{Remove?} (This is analogous to how $\isWFminDNE$ is a natural weakening of
    $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)

  \item At the same time, if the relation is finitely branching ($\FB$), then WFminDNE-
    implies WFacc- directly.

@@

% \item \apcomment{Fix with macros}. Since isWFminDNE- implies isWFseq-, the former also implies isWFminCor-.
%   But the converse fails: If a subset/predicate $P$ is $\lnot\lnot$-closed,
%   we cannot, in general, conclude it is coreductive.
  \item \apcomment{Delete? In general, the coreductive properties speak to the ability to choose
  a successor (strong existence).  This ability is central to being able to
  define sequences.  Therefore, sequential definitions generally require
some coreductive situation to be applicable.}

@@

We also did not include the inductive definition, Definition \ref{def:WFproperties}.\ref{def:WFind}.
  Logically, it is trivially equivalent to the accessible one, but its Agda formulation has a defect:
  the inductive well-foundedness property lands in the larger universe $\mathtt{Set}_1$, due to
  quantification over all predicates on $A$.  (This defect is also present in $\isWFmin$, $\isWFminDNE$, and \texttt{isWFcor}.)

@@

(The label $\isWFseqm$ refers to the name of the corresponding term in \texttt{Wellfoundedness.agda}.)
Constructively, this is not a useful hypothesis, as it does not provide any positive computational content and requires an explicit sequence to be defined.

@@

These two conditions are classically equivalent, but not constructively so.
Indeed, $\isWFseq$ certainly implies $\isWFseqm$, but the converse only holds classically.

@@

While classically equivalent to the previous one, see \cite[Exercise A.1.8]{Terese},
constructively we only have the implication $\isWFmin \then \isWFseq$.

@@

  _isWFacc : Set
  _isWFacc = ∀ (x : A) → x ∈ _-accessible
\end{verbatim}
}
Although the two notions are equivalent, the latter definition allows us to do induction on the proof of accessibility of an individual element $x \in A$.  This makes such proofs easier to write in Agda, thus we take
\verb|_isWFacc| as \emph{the} standard notion of well-foundedness for relations.

@@

For example, the archetype of a well-founded relation is the strict order \verb|<|
of natural numbers.  The usual induction readily validates $\isWFacc$ for \verb|<|.
  At the same time, the file \texttt{WFCounters.agda} shows that
asserting $\isWFmin$ for \verb|<| implies that every subset of the natural numbers is decidable.  Since this is clearly not provable in Agda without classical axioms,
this argument shows that the implication from $\isWFacc$ to $\isWFmin$ is not provable.

@@

On the other hand, $\isWFmin$ trivially implies that each $x \in A$ is weakly normalizing.
Indeed, given $x \in A$, consider the predicate $P_x(y) = x \mstep y$.
This predicate is non-empty, since the empty reduction yields a proof of $P_x(x)$.
Let $n$ be a $P_x$-minimal element.  Then $x \mstep n$.  Moreover, for any $y$,
If $n \rstep y$, then $\lnot(P_x(y))$.  But since reductions from $x$ are closed downward,
no such $y$ can exist, and hence $n$ is a normal form of $x$.

% \greyout{From the perspective of rewriting theory, it would seem wild indeed if
% $\SN$ did not imply $\WN$.
%
These questions motivated us to take a closer look at the issue of well-foundedness.
Our analysis revealed a number of additional related hypotheses, having a complex
network of relations.  %These are displayed in Figure \ref{f:wf}.
These are recorded in the file \verb|Wellfounded.agda|

>>>>

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks. Constructively, this is not a useful definition, as it does not provide any positive computational content and requires an explicit sequence to be defined.  This definition appears to be strictly weaker than
$\WFseqnn$, asserting that every sequence not-not contains a
non-decreasing index.

% \apcomment{Perhaps we should have a separate discussion between isWFseq- and isWFseq-not-not?}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  scale=0.6,
  every node/.style={transform shape},
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

% Row 1: Basic definitions
\node[box] (WFmin) {$\WFmin$};
\node[box, right=2.9cm of WFmin] (WFminDNE) {$\WFminDNE$};
\node[box, right=2.9cm of WFminDNE, line width = 0.5mm] (WFacc) {$\WFacc$};
\node[box, right=2.9cm of WFacc] (WFseq) {$\WFseq$};
\node[box, right=2.9cm of WFseq] (WFCor) {$\WFcor$};

% Row 2: Weak definitions
\node[box, below=2cm of WFmin] (WFminM) {$\WFminm$};
\node[box, below=2cm of WFminDNE] (WFminDNEM) {$\WFminDNEm$};
\node[box, below=2cm of WFacc] (WFaccM) {$\WFaccm$};
\node[box, below=2cm of WFseq] (WFseqM) {$\WFseqm$};
\node[box, below=2cm of WFCor] (WFminCor) {$\WFcorm$};

% Arrows within rows
\draw[arrow] (WFmin) -- (WFminDNE);
\draw[arrow, bend left=45] (WFmin) to (WFseq);

\draw[arrow] (WFminM) -- (WFminDNEM);
\draw[arrow, bend right=45] (WFminM) to (WFaccM);
\draw[arrow] (WFaccM) -- (WFseqM);
\draw[arrow, bend right=45] (WFaccM) to (WFminDNEM);
\draw[arrow, bend right=45] (WFminDNEM) to (WFminM);
\draw[arrow, bend right=45] (WFminDNEM) to (WFseqM);
\draw[arrow] (WFseqM) -- (WFminCor);

% Inter-row arrows
\draw[arrow] (WFmin) -- (WFminM);
\draw[arrow] (WFminDNE) -- (WFminDNEM);
\draw[arrow] (WFacc) -- (WFaccM);
\draw[arrow] (WFseq) -- (WFseqM);
\draw[arrow] (WFCor) -- (WFminCor);

% Additional labeled arrows
\draw[arrow] (WFminDNEM) to node[above, text=red] {$\FB$} (WFaccM);
\draw[arrow, bend right=45] (WFminCor) to node[above, text=red] {$\mpe$} (WFseqM);
\draw[arrow, bend right=45] (WFCor) to node[above, text=red] {$\accCor$} (WFacc);
\draw[arrow] (WFminDNE) to node[below, text=red]{$\accDNE$} (WFacc);
\draw[arrow, bend left = 45] (WFminDNE) to node[below, text=red]{$\mpe$} (WFseq);
\draw[arrow, bend right = 45] (WFminCor) to node[right, text=red]{$\corDNE$} (WFCor);
\draw[arrow, bend right = 45] (WFaccM) to node[right, text=red]{$\accDNE$} (WFacc);
\draw[arrow] (WFacc) to node[below, text=red]{$\rdec$} (WFseq);
\draw[arrow, bend left=45] (WFminCor) to node[below, text=red]{$\accCor$} (WFaccM);

\end{tikzpicture}
\caption{The logical relationships between notions of well-foundedness}
\label{fig:WF2}
\end{figure}

@@

{
\def\arraystretch{1.3}
\begin{figure}[h!]
\small
\begin{tabular}{@{}l l l @{}}
\toprule
\textbf{Property} & \textbf{Definition}  &\textbf{Agda Code}  \\
\midrule
$\rdec$   & $R$ is decidable
          & $\forall$ \verb|{x} {y}| $\to$ \verb|EM (R x y)|\\
$\FB$     & $R$ is finitely branching
          & $\mathtt{\forall (a : A) \to \Sigma [ xs \in List A ] (\forall b \to R a b \to b \in{}List\, xs)}$ \\
$\mpe$    & Sequences are $\lnot\lnot$-closed
          & $\forall\ \mathtt{(f : \nat \to A) \to \lnot\lnot Closed
          (\lambda x \to \Sigma [ k \in \nat ] (f\, k \equiv x))}$ \\
$\corDNE$ & Coreductives are $\lnot\lnot$-closed
          & $\mathtt{\forall (P : \mathcal{P} A) \to R}$-$\mathtt{coreductive\ P \to \lnot\lnot Closed\ P}$\\
$\accDNE$ & Accessibility is $\lnot\lnot$-closed
          & $\lnot\lnot$\verb|Closed (R -accessible)| \\
$\accCor$ & Accessibility is coreductive
          & \verb|R -coreductive (R -accessible)| \\
\bottomrule
\end{tabular}
\centering
\caption{Classical Properties Used in the Well-Foundedness Diagram}
\label{tab:cprop}
\end{figure}
}
\clearpage
% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.
Figure \ref{fig:WF} illustrates that the network of logical relationships between
constructive notions of well-foundedness is quite rich indeed.
Taking the $\lnot\lnot$-closures of
these notions cuts down some, but not all, of the complexity.
For example, the two minimality principles become equivalent.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the weak version of the sequential definition is unable to imply
the (weak) accessible one without an additional hypothesis $\accCor$,
which may be described as the constructive contrapositive of the definition of accessibility.
In the context of rewriting systems,
this condition asserts the existence of a function which maps any
element that is \emph{not} strongly normalizing to a one-step reduct of it having the same property.
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

Other observations that can be gleaned from the figure include the following.
\begin{enumerate}
  \item The implication from $\WFacc$ to $\WFseq$ requires the underlying relation to be decidable.
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item $\WFmin$ is a very strong assumption and implies $\WFseq$ unconditionally.
  The more reasonable $\WFminDNE$ requires the image of every sequence to be $\lnot\lnot$-closed
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (if one assumes that equality on $A$ is decidable).
  \item On the $\lnot\lnot$-side, $\WFminm$ and $\WFminDNEm$ are equivalent. Both are directly implied by $\WFaccm$,
    and all three directly imply $\WFseqm$.
  \item All definitions imply $\WFcorm$.  This weakest definition implies $\WFaccm$ assuming the
    ``constructive contrapositive of accessibility'', $\accCor$.
    %While this condition is implied by $\isWFseqm$, it does not appear to imply anything on its own.
  \item Assuming $\accCor$, $\WFacc$ can also be derived from $\WFcor$. % , the stronger version of $\WFcorm$.

    % \apcomment{Remove?} (This is analogous to how $\isWFminDNE$ is a natural weakening of
    % $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)

  \item At the same time, if the relation is finitely branching ($\FB$), then $\WFminDNEm$
    implies $WFaccm$ directly.

@@

% \item \apcomment{Fix with macros}. Since isWFminDNE- implies isWFseq-, the former also implies isWFminCor-.
%   But the converse fails: If a subset/predicate $P$ is $\lnot\lnot$-closed,
%   we cannot, in general, conclude it is coreductive.
%   \item \apcomment{Delete? In general, the coreductive properties speak to the ability to choose
%   a successor (strong existence).  This ability is central to being able to
%   define sequences.  Therefore, sequential definitions generally require
% some coreductive situation to be applicable.}

@@

We also did not include the inductive definition, Definition \ref{def:WFproperties}.\ref{def:WFind}.
  Logically, it is trivially equivalent to the accessible one, but its Agda formulation has a defect:
  the inductive well-foundedness property lands in the larger universe $\mathtt{Set}_1$, due to
  quantification over all predicates on $A$.  (This defect is also present in $\WFmin$, $\WFminDNE$, and \texttt{WFcor}.)

@@

(The label $\isWFseqm$ refers to the name of the corresponding term in \texttt{Wellfoundedness.agda}.)
Constructively, this is not a useful hypothesis, as it does not provide any positive computational content and requires an explicit sequence to be defined. \sacomment{This sentence added into where we discuss wfseq- as being the textbook definition. Can now be deleted here.}

@@

These two conditions are classically equivalent, but not constructively so.
Indeed, $\isWFseq$ certainly implies $\isWFseqm$, but the converse only holds classically. \sacomment{We say this through the figure and so can delete it?}

@@

While classically equivalent to the previous one, see \cite[Exercise A.1.8]{Terese},
constructively we only have the implication $\isWFmin \then \isWFseq$. \sacomment{Again, this is covered by our new definitions and figure and can be deleted.}

@@

  _isWFacc : Set
  _isWFacc = ∀ (x : A) → x ∈ _-accessible
\end{verbatim}
}
Although the two notions are equivalent, the latter definition allows us to do induction on the proof of accessibility of an individual element $x \in A$.  This makes such proofs easier to write in Agda, thus we take
\verb|_isWFacc| as \emph{the} standard notion of well-foundedness for relations.  \sacomment{We don't talk about which proof is easier to write in Agda, but I don't think we need to include this discussion. I suggest we delete.}

@@

For example, the archetype of a well-founded relation is the strict order \verb|<|
of natural numbers.  The usual induction readily validates $\WFacc$ for \verb|<|.
  At the same time, the file \texttt{WFCounters.agda} shows that
asserting $\WFmin$ for \verb|<| implies that every subset of the natural numbers is decidable.  Since this is clearly not provable in Agda without classical axioms,
this argument shows that the implication from $\WFacc$ to $\WFmin$ is not provable.

@@

On the other hand, $\WFmin$ trivially implies that each $x \in A$ is weakly normalizing.
Indeed, given $x \in A$, consider the predicate $P_x(y) = x \mstep y$.
This predicate is non-empty, since the empty reduction yields a proof of $P_x(x)$.
Let $n$ be a $P_x$-minimal element.  Then $x \mstep n$.  Moreover, for any $y$,
If $n \rstep y$, then $\lnot(P_x(y))$.  But since reductions from $x$ are closed downward,
no such $y$ can exist, and hence $n$ is a normal form of $x$.

% \greyout{From the perspective of rewriting theory, it would seem wild indeed if
% $\SN$ did not imply $\WN$.
%
These questions motivated us to take a closer look at the issue of well-foundedness.
Our analysis revealed a number of additional related hypotheses, having a complex
network of relations.  %These are displayed in Figure \ref{f:wf}.
These are recorded in the file \verb|Wellfounded.agda| /sacomment{Which file do we now want to refer to here. Or just refer to the entire subfolder?}
